A pipeline to analyse time-course gene expression data

### 2019-08-28 manuscript.tex
Deploy NelleV/2019timecourse-rnaseq-pipeline to github.com/NelleV/2019timecourse-rnaseq-pipeline.git:gh-pages

\begin{longtable}[]{@{}lll@{}}
\toprule
Name & Formula &\tabularnewline
\midrule
\endhead
timely & \(\text{lfc}(t)\) & Function of time\tabularnewline
sum & \(\sum_t lfc(t)\) & Sum of log fold change.\tabularnewline
abs\_sum & \(\sum_t \|lfc(t)\|\) & Always positive\tabularnewline
max & \(\max_t \|lfc(t)\|\) & Always positive\tabularnewline
min & \(\min_t \|lfc(t)\|\) & Always positive\tabularnewline
epicon & \$\$ & Captures overall strength of response and overall direction\tabularnewline
\bottomrule
\end{longtable}

\textbf{EAP: \texttt{epicon} is going to be weird for general use}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{log_fold_change_epicon =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
\NormalTok{    data, splines_model, timecourse_contrasts,  }\DataTypeTok{method=}\StringTok{"epicon"}\NormalTok{)}

@@

From the single measures of log-fold change and the p-value, we can now look
at the volcano plot. Here is an example of a volcano plot for the comparaison
of the control to the Kawasaki strain, using the EPICON log fold change
computation.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{pvalue =}\StringTok{ }\NormalTok{qvalues[, }\StringTok{"M-K"}\NormalTok{]}
\NormalTok{lfc_epicon =}\StringTok{ }\NormalTok{log_fold_change_epicon[, }\StringTok{"M-K"}\NormalTok{]}
\KeywordTok{names}\NormalTok{(lfc_epicon) =}\StringTok{ }\KeywordTok{row.names}\NormalTok{(log_fold_change_epicon)}

\KeywordTok{plot}\NormalTok{(lfc_epicon, }\OperatorTok{-}\KeywordTok{log10}\NormalTok{(pvalue), }\DataTypeTok{pch=}\DecValTok{20}\NormalTok{, }\DataTypeTok{main=}\StringTok{"Volcano plot"}\NormalTok{,}
     \DataTypeTok{xlim=}\KeywordTok{c}\NormalTok{(}\OperatorTok{-}\FloatTok{2.5}\NormalTok{, }\DecValTok{2}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

And here we visualize the genes with the largest absolute EPICON log-fold
change.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{top_DE_genes_lfc =}\StringTok{ }\KeywordTok{names}\NormalTok{(}
    \KeywordTok{sort}\NormalTok{(}\KeywordTok{abs}\NormalTok{(lfc_epicon),}
     \DataTypeTok{decreasing=}\OtherTok{TRUE}\NormalTok{)[}\DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{])}
\KeywordTok{plot_splines_data}\NormalTok{(data[top_DE_genes_lfc, ], splines_model,}
    \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{,}
    \DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{1.5}\NormalTok{,}\FloatTok{2.5}\NormalTok{,}\DecValTok{2}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

\CommentTok{# convert gene names}
\NormalTok{genes =}\StringTok{ }\KeywordTok{getBM}\NormalTok{(}\DataTypeTok{attributes=}\KeywordTok{c}\NormalTok{(}\StringTok{"ensembl_gene_id"}\NormalTok{, }\StringTok{"entrezgene_id"}\NormalTok{),}
        \DataTypeTok{filters=}\StringTok{"refseq_mrna"}\NormalTok{, }\DataTypeTok{values=}\NormalTok{genes,}
        \DataTypeTok{mart=}\NormalTok{ensembl)[}\StringTok{"entrezgene_id"}\NormalTok{]}
\NormalTok{genes =}\StringTok{ }\KeywordTok{as.vector}\NormalTok{(}\KeywordTok{unlist}\NormalTok{(genes))}
\NormalTok{pathways =}\StringTok{ }\KeywordTok{find_enriched_pathway}\NormalTok{(}
\NormalTok{    genes, }\DataTypeTok{species=}\StringTok{"mmu"}\NormalTok{,}
    \DataTypeTok{download_latest=}\OtherTok{TRUE}\NormalTok{)}\OperatorTok{$}\NormalTok{stastic}
\end{Highlighting}
\end{Shaded}

\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedleft}X>{\raggedleft}X}
\hline
  & Pathway & Percentage & Adj. p-value\\
\hline
05169 & Epstein-Barr virus infection & 37 & 0\\
\hline
04060 & Cytokine-cytokine receptor interaction & 31 & 0\\
\hline
04668 & TNF signaling pathway & 45 & 0\\
\hline
05164 & Influenza A & 38 & 0\\
\hline
04621 & NOD-like receptor signaling pathway & 34 & 0\\
\hline
05162 & Measles & 37 & 0\\
\hline
04630 & JAK-STAT signaling pathway & 33 & 0\\
\hline
05160 & Hepatitis C & 34 & 0\\
\hline
04064 & NF-kappa B signaling pathway & 39 & 0\\
\hline
04620 & Toll-like receptor signaling pathway & 40 & 0\\
\hline
\end{tabu}

>>>>

\begin{longtable}[]{@{}lll@{}}
\toprule
Name & Formula &\tabularnewline
\midrule
\endhead
timely & \(\text{lfc}(t)\) & Function of time\tabularnewline
sum & \(\sum_t lfc(t)\) & Sum of log fold change.\tabularnewline
abs\_sum & \(\sum_t \lvert lfc(t)\lvert\) & Always positive\tabularnewline
max & \(\max_t \lvert lfc(t) \lvert\) & Always positive\tabularnewline
min & \(\min_t \lvert lfc(t) \lvert\) & Always positive\tabularnewline
timecourse & See details below & Captures overall strength and direction\tabularnewline
\bottomrule
\end{longtable}

To capture the overall strength and direction of the response, we leverage the
timepoint by timepoint log-fold change \(\text{lfc}(t)\), and apply the
following formula:

\[\text{sign}\left(\frac{1}{T}\sum_{t = 1}^T lfc(t) \right) \times \left(\frac{1}{T}\sum_{t= 1}^T \lvert {lfc(t)} \lvert \right)\]

Note that when a gene is not consistently up or down-regulated the estimation
of the direction will not represent accurately the changes observed.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{log_fold_change_timecourse =}\StringTok{ }\KeywordTok{estimate_log_fold_change}\NormalTok{(}
\NormalTok{    data, splines_model, timecourse_contrasts,  }\DataTypeTok{method=}\StringTok{"timecourse"}\NormalTok{)}

@@

From the single measures of log-fold change and the p-value, we can now look
at the volcano plot. Here is an example of a volcano plot for the comparaison
of the control to the Kawasaki strain, using the timecourse log fold change
computation.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{pvalue =}\StringTok{ }\NormalTok{qvalues[, }\StringTok{"M-K"}\NormalTok{]}
\NormalTok{lfc_timecourse =}\StringTok{ }\NormalTok{log_fold_change_timecourse[, }\StringTok{"M-K"}\NormalTok{]}
\KeywordTok{names}\NormalTok{(lfc_timecourse) =}\StringTok{ }\KeywordTok{row.names}\NormalTok{(log_fold_change_timecourse)}

\KeywordTok{plot}\NormalTok{(lfc_timecourse, }\OperatorTok{-}\KeywordTok{log10}\NormalTok{(pvalue), }\DataTypeTok{pch=}\DecValTok{20}\NormalTok{, }\DataTypeTok{main=}\StringTok{"Volcano plot"}\NormalTok{,}
     \DataTypeTok{xlim=}\KeywordTok{c}\NormalTok{(}\OperatorTok{-}\FloatTok{2.5}\NormalTok{, }\DecValTok{2}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

And here we visualize the genes with the largest absolute timecourse log-fold
change.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{top_DE_genes_lfc =}\StringTok{ }\KeywordTok{names}\NormalTok{(}
    \KeywordTok{sort}\NormalTok{(}\KeywordTok{abs}\NormalTok{(lfc_timecourse),}
     \DataTypeTok{decreasing=}\OtherTok{TRUE}\NormalTok{)[}\DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{])}
\KeywordTok{plot_splines_data}\NormalTok{(data[top_DE_genes_lfc, ], splines_model,}
    \DataTypeTok{colors=}\NormalTok{ann_colors}\OperatorTok{$}\NormalTok{Group, }\DataTypeTok{smooth=}\OtherTok{TRUE}\NormalTok{,}
    \DataTypeTok{mar=}\KeywordTok{c}\NormalTok{(}\FloatTok{1.5}\NormalTok{,}\FloatTok{2.5}\NormalTok{,}\DecValTok{2}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

\CommentTok{# convert gene names}
\NormalTok{genes =}\StringTok{ }\KeywordTok{getBM}\NormalTok{(}\DataTypeTok{attributes=}\KeywordTok{c}\NormalTok{(}\StringTok{"ensembl_gene_id"}\NormalTok{, }\StringTok{"entrezgene_id"}\NormalTok{),}
        \DataTypeTok{filters=}\StringTok{"refseq_mrna"}\NormalTok{, }\DataTypeTok{values=}\NormalTok{genes,}
        \DataTypeTok{mart=}\NormalTok{ensembl)[}\StringTok{"entrezgene_id"}\NormalTok{]}
\NormalTok{genes =}\StringTok{ }\KeywordTok{as.vector}\NormalTok{(}\KeywordTok{unlist}\NormalTok{(genes))}
\NormalTok{pathways =}\StringTok{ }\KeywordTok{find_enriched_pathway}\NormalTok{(}
\NormalTok{    genes, }\DataTypeTok{species=}\StringTok{"mmu"}\NormalTok{,}
    \DataTypeTok{download_latest=}\OtherTok{FALSE}\NormalTok{)}\OperatorTok{$}\NormalTok{stastic}
\end{Highlighting}
\end{Shaded}

\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedleft}X>{\raggedleft}X}
\hline
  & Pathway & Percentage & Adj. p-value\\
\hline
04060 & Cytokine-cytokine receptor interaction & 34 & 0\\
\hline
04620 & Toll-like receptor signaling pathway & 40 & 0\\
\hline
04630 & Jak-STAT signaling pathway & 33 & 0\\
\hline
03050 & Proteasome & 53 & 0\\
\hline
04621 & NOD-like receptor signaling pathway & 47 & 0\\
\hline
04380 & Osteoclast differentiation & 34 & 0\\
\hline
04623 & Cytosolic DNA-sensing pathway & 45 & 0\\
\hline
04210 & Apoptosis & 37 & 0\\
\hline
04622 & RIG-I-like receptor signaling pathway & 39 & 0\\
\hline
05160 & Hepatitis C & 29 & 0\\
\hline
\end{tabu}
